---
title: "01_HEU_HUU_metabolomics"
format: html
editor: visual
---

## Notebook Setup

```{r}

#| label: setup
#| warning: false

setwd("/Volumes/T7/Li Lab/github/2025-08-04_HEU_analysis/immune_profile_analysis/heu_metabolomics_manuscript/")

this_date <- Sys.Date()

output_dir <- glue::glue("/Volumes/T7/Li Lab/github/2025-08-04_HEU_analysis/immune_profile_analysis/heu_metabolomics_manuscript/{this_date}_processed_data/")
dir.create(output_dir)

results_dir <- glue::glue("/Volumes/T7/Li Lab/github/2025-08-04_HEU_analysis/immune_profile_analysis/heu_metabolomics_manuscript/{this_date}_results/")
dir.create(results_dir)

met_data_dir <- ("/Volumes/T7/Li Lab/Projects/HEU/OLD/2025-12-14_HEU_Metabolomics_V3/heu_data/")

library(tidyverse)
library(matrixTests)
library(ggrepel)
library(ComplexHeatmap)
library(showtext)

font_add("Arial", "/System/Library/Fonts/Supplemental/Arial.ttf")
showtext_opts(dpi = 300)
showtext_auto(enable = TRUE)
```

## Process Metabolomics Data

### Processing the data

```{r}

modes <- c("HILIC_neg", "HILIC_pos", "RP_neg", "RP_pos") 

for(m in modes){
  
  cat(m, "data processing beginning...")
  
  this_met_output_dir <- glue::glue("{output_dir}{results_date}_{m}/")
  dir.create(this_met_output_dir)
  
  ion_mode <- ifelse(grepl("pos", m), "pos", "neg")
  
  ft_tbl_path <- glue::glue("{met_data_dir}/12_03_2025_feature_tables/{m}_final_table.tsv")

  ft_tbl <- readr::read_tsv(ft_tbl_path)
  
  if(m == "RP_neg"){
    
    ft_tbl_path_full <- glue::glue("{met_data_dir}/12_03_2025_final_feature_tables/{m}_final_table.tsv")
    
    ft_tbl_final <- readr::read_tsv(ft_tbl_path_full)
    
    efa_ft <- "F29911"
    hyroxy_ft <- "F21642"
    
    if(!efa_ft %in% ft_tbl$id_number){
      
      ft_tbl <- dplyr::bind_rows(ft_tbl,
                                 ft_tbl_final[ft_tbl_final$id_number == efa_ft,])
      
    }
    
    if(!hyroxy_ft %in% ft_tbl$id_number){
      
      ft_tbl <- dplyr::bind_rows(ft_tbl,
                                 ft_tbl_final[ft_tbl_final$id_number == hyroxy_ft,])
      
    }
    
  }
  
  ## format annotations
  annot_tbl_path <- glue::glue("{met_data_dir}12_03_2025_annotations/{m}_deduped.tsv")
  
  annot_tbl <- readr::read_delim(annot_tbl_path, delim = ",")
  
  # # get kipus with stable isotopes from JSON
  kp_path <- glue::glue("{met_data_dir}/12_03_2025_empirical_compounds/{m}_empirical_compounds.csv")
  
  kp_df <- read.csv(kp_path) %>%
    group_by(kp_id) %>%
    mutate(
      has_stable_isotope = case_when(ion_mode == "pos" ~  max(grepl("13C/12C,M+H+", ion_relation, fixed = TRUE)),
                                     ion_mode == "neg" ~ max(grepl("13C/12C,M-H-", ion_relation, fixed = TRUE))),

      has_M0 = case_when(ion_mode == "pos" ~ max(grepl("M0,M+H+", ion_relation, fixed = TRUE)),
                         ion_mode == "neg" ~ max(grepl("M0,M-H-", ion_relation, fixed = TRUE)))
    )

  # confident C12s
  kp_df_df_filt <- kp_df %>%
    filter(has_stable_isotope & has_M0) %>%
    filter(isotope == "M0")

  if(ion_mode == "pos"){
    kp_df_df_filt <- kp_df_df_filt %>%
      filter(ion_relation == "M0,M+H+")
  }else{
    kp_df_df_filt <- kp_df_df_filt %>%
      filter(ion_relation == "M0,M-H-")
  }
  
  # add efavirenz
  efz_fts <- annot_tbl %>%
    filter(grepl("efavirenz", tolower(annotation)))
  
  kp_df_df_filt <- kp_df_df_filt %>%
    dplyr::bind_rows(.,
                     kp_df %>%
    filter(id %in% efz_fts$id_number)) %>%
    dplyr::left_join(., annot_tbl,
                     by = join_by(id == id_number)) %>%
    # Could be duplicates
    distinct()
  
  cat(nrow(kp_df_df_filt), " khipus in", m)

  write.csv(kp_df_df_filt,
          glue::glue("{this_met_output_dir}{m}_kp_df_anot.csv"))

  meta_data_path <- glue::glue("{met_data_dir}12_03_2025_mapping_files/sample_mapping_{m}_HEU_all_plasma.json")

  meta_json <- readLines(meta_data_path, warn = FALSE)
  meta_json <- gsub("\\bNaN\\b", "null", meta_json)  # replace NaN with null
  meta_json <- jsonlite::fromJSON(paste(meta_json, collapse = "\n"))

  meta_df <- tibble(sample_name = names(meta_json),
                    inner = meta_json) %>%
    unnest_wider(inner) %>%
    mutate(origin = ifelse(grepl("B", Sample_ID), "Cord", "Maternal"),
           viral_load__M = unlist(lapply(viral_load__M, function(y) if (is.null(y)) NA else y), use.names = TRUE))

  pref_ft_tbl_sliced <- ft_tbl %>%
    dplyr::select(c(1, 12:ncol(ft_tbl))) %>%
    pivot_longer(-1) %>%
    pivot_wider(names_from = 1, values_from = value) %>%
    tibble::column_to_rownames(var = names(.)[1]) %>%
    as.data.frame()

  # drop rows not in sample_metadata
  filtered_df_tbl <- pref_ft_tbl_sliced[rownames(pref_ft_tbl_sliced) %in% meta_df$sample_name, ]

  smpl.inx <- match(meta_df$sample_name, rownames(filtered_df_tbl))

  filtered_df_tbl <- filtered_df_tbl[smpl.inx,]

  filtered_df_tbl_t <- data.table::transpose(filtered_df_tbl)
  rownames(filtered_df_tbl_t) <- colnames(filtered_df_tbl)
  colnames(filtered_df_tbl_t) <- rownames(filtered_df_tbl)

  # perform log transformation - already done in pcpfm
  #log_ft_tbl <- log(filtered_df_tbl_t + 1)

  # keep features that are high-confidence?
  #log_ft_tbl_m0 <- log_ft_tbl[rownames(log_ft_tbl) %in% kp_df_df_filt$id, ]

  # rows features
  # columns samples
  write.csv(filtered_df_tbl_t,
            glue::glue("{this_met_output_dir}{m}_log_feature_table.csv"),
            row.names = TRUE,
            quote = TRUE,
            col.names = NA)

  write.csv(meta_df, glue::glue("{this_met_output_dir}{m}_metadata.csv"))

  write.csv(annot_tbl %>%
              filter(id_number %in% rownames(filtered_df_tbl_t)) %>%
              select(-'...1'), 
            glue::glue("{this_met_output_dir}{m}_annotations.csv"))
  
  cat(m, "data processing finished!")
    
}

```

## General Analysis

### Functions for analysis

```{r}

plot_boxplot_series <- function(top_volcano_ft,
                                n_top = 5,
                                this_ft_tbl,
                                these_samples,
                                kp_df,
                                m,
                                who,
                                bp_dir,
                                analysis = "volcano",
                                this_feat = NA,
                                series_name = NA,
                                nrow_facet = 1,
                                width = 7,
                                height = 2.5){

  if(analysis == "volcano"){
    
    # care about efavirenz and top n features based on p-values
    top_features <- top_volcano_ft %>%
    dplyr::left_join(., kp_df %>%
                       select(-annotation),
                     by = join_by(feature_id == id)) %>%
    # focus on high-confidence features
    dplyr::filter(!is.na(mz) | grepl("efavirenz", tolower(annotation))) %>%
    mutate(plot_name = gsub("_.*", "", annotation),
           plot_name = glue::glue("{stringr::str_trunc(plot_name, width = 20)}\n{round(mz, 4)}@{rtime}\n{m}")) %>%
    slice_head(n = n_top)
    
    print(head(top_features))
    
  }else if(analysis == "single"){
    
    # plot for single feature
    top_features <- top_volcano_ft %>%
      dplyr::filter(feature_id %in% this_feat) %>%
      dplyr::left_join(., kp_df %>%
                       select(-annotation),
                     by = join_by(feature_id == id)) %>%
    mutate(plot_name = gsub("_.*", "", annotation),
           plot_name = glue::glue("{stringr::str_trunc(plot_name, width = 20)}\n{round(mz, 4)}@{rtime}\n{m}")) 
    
  }else if(analysis == "series"){
    
    top_features <- top_volcano_ft %>%
    dplyr::left_join(., kp_df %>%
                       select(-annotation) %>%
                       tibble::rownames_to_column(., "feature_id")) %>%
    mutate(plot_name = gsub("_.*", "", annotation),
           plot_name = glue::glue("{stringr::str_trunc(plot_name, width = 20)}\n{round(mz, 4)}@{rtime}\n{m}")) %>%
    slice_head(n = n_top)
    
    print(head(top_features))
    
  }
  
  top_ft_df <- this_ft_tbl %>%
    as.data.frame() %>%
    select(all_of(top_features$feature_id)) %>%
    tibble::rownames_to_column(.) %>%
    dplyr::rename(sample_name = rowname) %>%
    filter(sample_name %in% these_samples$sample_name) %>%
    dplyr::left_join(., these_samples %>%
                       select(sample_name, status)) %>%
    mutate(status = factor(status, levels = c("HUU", "HEU"))) %>%
    pivot_longer(., cols = all_of(top_features$feature_id), names_to = "feature_id") %>%
    dplyr::left_join(., top_features %>%
                       select(feature_id, plot_name)) %>%
    mutate(plot_name = factor(plot_name,
                              levels = unique(top_features$plot_name)))
  
  p <- ggplot(top_ft_df, 
              aes(x = status, 
                  y = value)) +
    geom_boxplot(alpha = 0.85,
                 outlier.shape = NA,
                 aes(fill = status)) +
    geom_jitter(aes(fill = status),
                width = 0.15, 
                alpha = 0.6, 
                size = 1.75,
                shape = 21, 
                stroke = 0.5,
                color = "black") +
    scale_fill_manual(values = c("HUU" = "#1A85FF", 
                                 "HEU" = "#D41159"),
                      guide = "none") +
    theme_classic() +
    labs(x = "HIV Exposure Status",
         y = "Relative Intensity") +
    theme(
      text = element_text(family = "Arial"),
      strip.background = element_blank(),
      panel.spacing = unit(0.5, "lines"))
  
  if(analysis %in% c("volcano")){
    p <- p + 
      facet_wrap(. ~ plot_name, nrow = nrow_facet, scales = "free_y") +
      theme(
        axis.text = element_text(size = 9),
      axis.title.x = element_text(size = 9, margin = margin(t = 2.5)), # reduce space above x-axis title
      axis.title.y = element_text(size = 9, margin = margin(r = 2.5)), # reduce space right of y-axis title
      plot.title = element_text(size = 9.5, hjust = 0.5),
      strip.text = element_text(size = 8, margin = margin(b = 1, t = 1)),
      ) +
    ggtitle(glue::glue("{m}: HUU vs HEU {who} Plasma"))
    
    ggsave(glue::glue("{bp_dir}{Sys.Date()}_{m}_{who}_box_plot.png"),
         width = width,
         height = height)
    
  }else if(analysis %in% c("series")){
    p <- p + 
      facet_wrap(. ~ plot_name, nrow = nrow_facet, scales = "free_y") +
      theme(
        axis.text = element_text(size = 9),
      axis.title.x = element_text(size = 9, margin = margin(t = 2.5)), # reduce space above x-axis title
      axis.title.y = element_text(size = 9, margin = margin(r = 2.5)), # reduce space right of y-axis title
      plot.title = element_text(size = 9, hjust = 0.5),
      strip.text = element_text(size = 8, margin = margin(b = 1, t = 1)),
      ) +
    ggtitle(glue::glue("{series_name}\n{m}: HUU vs HEU {who} Plasma"))
    
    ggsave(glue::glue("{bp_dir}{Sys.Date()}_{m}_{who}_{series_name}_box_plot.png"),
         width = width,
         height = height)
    
  }else{
    
    p <- p + 
      theme(
        axis.text = element_text(size = 6),
      axis.title.x = element_text(size = 6, margin = margin(t = 2.5)), # reduce space above x-axis title
      axis.title.y = element_text(size = 6, margin = margin(r = 2.5)), # reduce space right of y-axis title
      plot.title = element_text(size = 6.5, hjust = 0.5),
      strip.text = element_text(size = 5, margin = margin(b = 1, t = 1)),
      ) +
    ggtitle(top_ft_df$plot_name)
    
    ggsave(glue::glue("{bp_dir}{Sys.Date()}_{m}_{who}_{janitor::make_clean_names(top_features$annotation)}_{this_feat}_box_plot.png"),
         width = 1.5,
         height = 2,
         dpi = 300)
  }
}

plot_heatmap <- function(paired_ttest_res,
                         this_ft_tbl,
                         this_metadata,
                         m,
                         heatmap_dir){
  
  sig_fts <- paired_ttest_res %>%
    tibble::rownames_to_column(., "feature_id") %>%
    filter(p.adjusted <= 0.05) %>%
    pull(feature_id)
  
  sig_fts_cleaned <- gsub("_delta", "", sig_fts)
  
  ft_tbl_scaled <- this_ft_tbl %>%
    as.data.frame() %>%
    select(all_of(sig_fts_cleaned)) %>%
    scale() %>%
    as.matrix() %>%
    t()
  
  col_annots <- this_metadata %>%
    select(status,
           sex__B,
           origin,
           Batch_no) %>%
    mutate(
      sex__B = ifelse(origin == "Maternal", "Maternal", sex__B),
      cord_sex = ifelse(sex__B == "Male", "Cord M", 
                        ifelse(sex__B == "Female", "Cord F", sex__B))
    )
  
  column_ha <- HeatmapAnnotation(Batch = col_annots$Batch_no,
                                 Origin = col_annots$cord_sex,
                                 Status = col_annots$status,
                                 col = list(
                                   Origin = c("Cord F" = "#CC79A7", 
                                              "Cord M" = "#56B4E9", 
                                              "Maternal" = "#F0E442"),
                                   Status = c("HUU" = "#1A85FF", 
                                              "HEU" = "#D41159")
                                 ),
                                 # annotation track titles
                                 annotation_name_gp = gpar(fontfamily = "Arial", 
                                                           fontsize = 14, 
                                                           fontface = "plain"),
                                 # annotation legends
                                 annotation_legend_param = list(
                                   Batch  = list(ncol = 1, 
                                                 title_gp = gpar(fontfamily = "Arial", 
                                                                 fontsize = 14, 
                                                                 fontface = "plain"),
                                                 labels_gp = gpar(fontfamily = "Arial", 
                                                                  fontsize = 14, 
                                                                  fontface = "plain")),
                                   Origin = list(ncol = 1, 
                                                 title_gp = gpar(fontfamily = "Arial", 
                                                                 fontsize = 14, 
                                                                 fontface = "plain"),
                                                 labels_gp = gpar(fontfamily = "Arial", 
                                                                  fontsize = 14, 
                                                                  fontface = "plain")),
                                   Status = list(ncol = 1, 
                                                 title_gp = gpar(fontfamily = "Arial", 
                                                                 fontsize = 14, 
                                                                 fontface = "plain"),
                                                 labels_gp = gpar(fontfamily = "Arial", 
                                                                  fontsize = 14, 
                                                                  fontface = "plain"))))
  
  ht <- Heatmap(ft_tbl_scaled, 
                column_title = glue::glue("Clustermap q < 0.05 (BH) {m}"),
                column_title_gp = gpar(fontfamily = "Arial", 
                                       fontsize = 16, 
                                       fontface = "plain"),
                use_raster = FALSE,
                show_row_names = FALSE,
                show_column_names = FALSE,
                top_annotation = column_ha,
                name = "Z-Score (Feature)",
                # main heatmap legend
                heatmap_legend_param = list(
                  title_gp = gpar(fontfamily = "Arial", 
                                  fontsize = 14, 
                                  fontface = "plain"),
                  labels_gp = gpar(fontfamily = "Arial", 
                                   fontsize = 14,
                                   fontface = "plain")
                ))  
  
  pdf(glue::glue("{heatmap_dir}{Sys.Date()}_{m}_all_samples_heatmap.pdf"), 
      width = 8, 
      height = 6)
  
  showtext_begin()
  draw(ht)        
  showtext_end()
  
  dev.off() 
  
}

plot_pca <- function(m,
                     this_ft_tbl,
                     this_metadata,
                     pca_dir){
  
  pc <- stats::prcomp(na.omit(this_ft_tbl),
                      center = TRUE,
                      scale = TRUE)
  
  pc$percentVar <- pc$sdev^2/sum(pc$sdev^2)
  
  pca_res <- as.data.frame(pc$x)
  pca_res$origin <- as.factor(this_metadata$origin)
  pca_res$status <- as.factor(this_metadata$status)
  pca_res$group <- as.factor(this_metadata$plot_group)
  
  p <- ggplot2::ggplot(data = pca_res,
                       ggplot2::aes(x = PC1,
                                    y = PC2,
                                    color = status)) +
    ggplot2::geom_jitter(aes(shape = origin),
                         size = 5,
                         alpha = 0.75) +
    ggplot2::theme_classic() +
    scale_color_manual(values = c("HUU" = "#1A85FF", 
                                  "HEU" = "#D41159")) +
    ggplot2::scale_shape_manual(values = c(16, 15)) +
    ggplot2::labs(x = glue::glue("PC1: ", round(pc$percentVar[1]*100,1), "%"),
                  y = glue::glue("PC2: ", round(pc$percentVar[2]*100,1), "%")) +
    ggplot2::theme(text = ggplot2::element_text(size = 12,
                                                family = "Arial"),
                   axis.text = ggplot2::element_text(size = 12),
                   axis.title = ggplot2::element_text(size = 14),
                   panel.grid = element_blank()) +
    stat_ellipse(aes(group = group)) +
    #tune::coord_obs_pred() +
    labs(shape = "Origin",
         color = "HIV Status",
         title = glue::glue("PCA of {m}"))
  
  ggplot2::ggsave(glue::glue("{pca_dir}{Sys.Date()}_{m}_pca_plot.png"),
                  width = 6,
                  height = 4.5)
  
}

# paired t-tests

run_paired_ttests <- function(m,
                              this_ft_tbl,
                              this_metadata,
                              output_dir){
  
  all_features <- colnames(this_ft_tbl)
  
  this_ft_tbl_df <- this_ft_tbl %>%
    as.data.frame() %>%
    mutate(sample_name = rownames(this_ft_tbl))
  
  all_data_df <- dplyr::left_join(this_metadata %>%
                                    select(sample_name, Sample_ID, status, origin),
                                  this_ft_tbl_df)
  
  all_data_df_delta <- all_data_df %>%
    mutate(dyad_id = gsub("_.*", "", Sample_ID)) %>%
    group_by(dyad_id) %>%
    mutate(n = n()) %>%
    filter(n == 2) %>%
    arrange(Sample_ID, origin) %>%
    mutate(across(all_of(all_features), ~ (.x[1] - .x[2]), .names = "{.col}_delta")) %>%
    select(-c(all_of(all_features))) %>%
    distinct()
  
  tt_df <- all_data_df_delta %>%
    select(dyad_id, status, all_of(paste0(all_features, "_delta"))) %>%
    distinct() %>%
    ungroup()
  
  tt_df_huu <- tt_df %>%
    dplyr::filter(status == "HUU") %>%
    select(all_of(paste0(all_features, "_delta")))
  
  tt_df_heu <- tt_df %>%
    dplyr::filter(status == "HEU") %>%
    select(all_of(paste0(all_features, "_delta")))
  
  tt_res <- matrixTests::col_t_welch(tt_df_huu, 
                                     tt_df_heu) 
  
  tt_res <- tt_res %>%
    filter(!obs.x == 0) %>%
    mutate(p.adjusted = p.adjust(pvalue, method = "BH"))
  
  write.csv(tt_res,
            glue::glue("{output_dir}{Sys.Date()}_{m}_paired_ttest_results_delta_dyad.csv"))
  

  return(tt_res)
  
}

#' Perform t-test
#' per feature between two given groups.
#'
#' @param peak_table Data frame with samples as rows and features as columns.
#' @param metadata Data frame with each row containing
#' metadata for each sample.
#' @param group1 Character, exact name of the first group.
#' @param group2 Character, exact name of the second group.
#' @param prefix Character, prefix to append results.
#' @param output_dir File path to save results.
#' @import matrixTests
#' @export
perform_ttest <- function(peak_table = NULL,
                          metadata = NULL,
                          group_var = NA,
                          group1 = "",
                          group2 = "",
                          prefix = "",
                          output_dir = "",
                          type = "default"){

  if(missing(peak_table) | missing(metadata) | missing(group_var) | missing(group1) | missing(group2)){
    cli::cli_abort("Missing peak table and/or metadata and/or grouping variable!")
  }

  g1inx <- which(metadata[[group_var]] == group1)
  g2inx <- which(metadata[[group_var]] == group2)

  if(!group_var %in% colnames(metadata)){
    cli::cli_abort("Grouping variable not found in metadata!")
  }

  peak1 <- peak_table[,g1inx, drop = FALSE]
  peak2 <- peak_table[,g2inx, drop = FALSE]

  if(any(ncol(peak1) == 0 | ncol(peak2) == 0)){
    cli::cli_abort("Invalid group sizes! At least 1 group has no corresponding samples!")
  }

  if(ncol(peak1) == 1){
    peak1 <- peak1 %>% as.data.frame() %>% dplyr::bind_cols((.))
  }

  if(ncol(peak2) == 1){
    peak2 <- peak2 %>% as.data.frame() %>% dplyr::bind_cols((.))
  }

  if(type == "default"){
    # perform two-sample t-test
    tt_res <- matrixTests::row_t_welch(peak1, peak2)
  }else if(type == "paired"){
    tt_res <- matrixTests::row_t_paired(peak1, peak2)
  }

  tt_res$pvalue_adjusted <- p.adjust(tt_res$pvalue, "BH")

  write.csv(tt_res,
            glue::glue("{output_dir}{prefix}_ttest_results.csv"))

  return(tt_res)

}

# Assumes peak table is already log fold-changed!
perform_foldchange <- function(peak_table = pk_tbl,
                               metadata = meta_tbl,
                               group_var = "group_ko_time",
                               group1 = group1,
                               group2 = group2,
                               prefix = glue::glue("{this_mode}_{group1}_{group2}"),
                               output_dir = results_dir){
  
  g1inx <- which(metadata[[group_var]] == group1)
  g2inx <- which(metadata[[group_var]] == group2)
  
  features <- rownames(peak_table)
  
  fc_res <- lapply(features, function(x){
    
    feats_g1 <- peak_table[which(rownames(peak_table) == x), g1inx, drop = FALSE] %>% unlist()
    feats_g2 <- peak_table[which(rownames(peak_table) == x), g2inx, drop = FALSE] %>% unlist()
    
    mean_g1 <- mean(feats_g1, na.rm = TRUE)
    mean_g2 <- mean(feats_g2, na.rm = TRUE)
    log2fc <- mean_g2 - mean_g1 
    
    res <- data.frame(log2fc = log2fc)
    rownames(res) <- x
    
    res
    
  })
  
  fc_res <- fc_res %>%
    bind_rows()
  
  write.csv(fc_res,
            glue::glue("{output_dir}{prefix}_foldchange_results.csv"))
  
  return(fc_res)
  
}

plot_volcano <- function(m,
                         tt_res = tt_df,
                         pval_thresh = 0.05,
                         fc_res = fc_df,
                         fc_thresh = 1.5,
                         prefix = glue::glue("{Sys.Date()}_{group1}_{group2}_{who}"),
                         output_dir = results_dir,
                         pvalue_type = "adjusted",
                         width = 6,
                         height = 5,
                         labels = TRUE,
                         anot_df = this_annots,
                         highlight_metabolites = TRUE,
                         metabolite_pattern = NA){
  
  all_res_df <- tt_df %>%
    mutate(tibble::rownames_to_column(., "feature_id")) %>%
    dplyr::left_join(fc_df %>%
                       mutate(tibble::rownames_to_column(., "feature_id"))) %>%
    mutate(
      important = ifelse(pvalue_adjusted < pval_thresh & log2fc > log2(fc_thresh), "TT + FC Up",
                         ifelse(pvalue_adjusted < pval_thresh & log2fc < -log2(fc_thresh), "TT + FC Down", "No Change")),
      important = ifelse((pvalue_adjusted < pval_thresh & important == "No Change"), "TT Only", important),
      important = factor(important))
  
  if(highlight_metabolites){
    all_res_df <- all_res_df %>%
      dplyr::left_join(., anot_df) %>%
      mutate(
        highlighted_met = ifelse(grepl(metabolite_pattern, tolower(annotation)), annotation, NA),
        markers_plot = ifelse(important != "No Change", highlighted_met, NA)
      )
  }else{
    all_res_df <- all_res_df %>%
      mutate(markers_plot = NA)
  }

  n_imp <- all_res_df %>%
    filter(important != "No Change") %>%
    count() %>%
    unlist()
  
  cat(glue::glue("{n_imp} significant features!"))
  
  color_pal <- c("red", "blue", "#D8BB64", "grey")
  names(color_pal) <- c("TT + FC Up", "TT + FC Down", "TT Only", "No Change")
  
  max_x <- max(abs(all_res_df$log2fc), na.rm = TRUE)
  
  if(max_x < 0.5){
    max_x <- 0.5
  } else{
    max_x <- ceiling(max_x)
  }
  
  p <- ggplot(all_res_df, 
              aes(x = log2fc, 
                  y = -log10(pvalue_adjusted),
                  color = important,
                  label = markers_plot)) +
    geom_point(size = 2,
               alpha = 0.75) +
    geom_hline(yintercept = -log10(pval_thresh), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = c(-log2(fc_thresh), log2(fc_thresh)), linetype = "dashed", color = "grey40") +
    theme_classic() +
    scale_color_manual(values = color_pal) +
    labs(
      x = "Log2 Fold Change",
      y = "-Log10 (BH adjusted TT p-value)",
      color = "Significance"
    ) +
    theme(
      text = element_text(family = "Arial"),
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12),
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      strip.text = element_text(size = 11),
      panel.grid = element_blank()) +
    coord_cartesian(xlim = c(-max_x, max_x)) +
    ggtitle(glue::glue("{m} HUU vs HEU {who} - BH < {pval_thresh}")) +
  geom_text_repel(
    na.rm = TRUE,
    size = 3,
    color = "black",
    box.padding = 0.3,
    point.padding = 0.5,
    max.overlaps = Inf,
    min.segment.length = 0.1
  )
  
  ggsave(glue::glue("{volcano_dir}{prefix}_volcano_plot.png"),
         width = 5,
         height = 4)
  
  write.csv(all_res_df, glue::glue("{volcano_dir}{prefix}_volcano_df.csv"))
  
  supp_table <- all_res_df %>%
    rename(
      n_HUU = obs.x,
      n_HEU = obs.y,
      mean_HUU = mean.x,
      mean_HEU = mean.y,
      Log2FC = log2fc,
      TT_pvalue = pvalue,
      TT_adjpvalue = pvalue_adjusted
    ) %>%
    mutate(mean_HUU = signif(mean_HUU, 2),
           mean_HEU = signif(mean_HEU, 2),
           Log2FC = signif(Log2FC, 2),
           TT_pvalue = signif(TT_pvalue, 2),
           TT_adjpvalue = signif(TT_adjpvalue, 2)) %>%
    select(feature_id, n_HUU, n_HEU, mean_HUU, mean_HEU, 
           Log2FC, TT_pvalue, TT_adjpvalue) %>%
    arrange(TT_adjpvalue)
  
  write.csv(supp_table, glue::glue("{volcano_dir}{prefix}_supp_volcano_table_df.csv"))
  
}
```

### Analyze metabolomics data

```{r}

modes <- c("HILIC_neg", "HILIC_pos", "RP_neg", "RP_pos") 

figure1_dir <- glue::glue("{results_dir}figure1/")
dir.create(figure1_dir)

pca_dir <- glue::glue("{figure1_dir}pca/")
dir.create(pca_dir)

ttest_dir <- glue::glue("{figure1_dir}ttest/")
dir.create(ttest_dir)

fc_dir <- glue::glue("{figure1_dir}fc/")
dir.create(fc_dir)

volcano_dir <- glue::glue("{figure1_dir}volcano/")
dir.create(volcano_dir)

heatmap_dir <- glue::glue("{figure1_dir}heatmap/")
dir.create(heatmap_dir)

bp_dir <- glue::glue("{figure1_dir}boxplots/")
dir.create(bp_dir)

group1 <- "HUU"
group2 <- "HEU"
group_var = "status"

results_date = "2025-12-16"

for(m in modes){
  
  this_met_output_dir <- glue::glue("{output_dir}{results_date}_{m}/")
  
  this_ft_tbl <- list.files(this_met_output_dir,
                            pattern = "feature_table",
                            full.names = TRUE) %>%
    read.csv(., row.names = 1) %>%
    t()
  
  this_metadata <- list.files(this_met_output_dir,
                              pattern = "metadata",
                              full.names = TRUE) %>%
    read.csv(., row.names = 1) %>%
    mutate(who = ifelse(origin == "Maternal", "Maternal", "Cord"),
           plot_group = paste0(who, "_", status))
  
  this_annots <- list.files(this_met_output_dir,
                            pattern = "annotations",
                            full.names = TRUE) %>%
    read.csv(., row.names = 1) %>%
    rename(feature_id = id_number)
  
  kp_path <- glue::glue("{met_data_dir}/12_03_2025_empirical_compounds/{this_m}_empirical_compounds.csv")
    
  kp_df <- read.csv(kp_path) %>%
    mutate(annotation = NA)
  
  # M0 only
  #kp_df <- glue::glue("{this_met_output_dir}{m}_kp_df_anot.csv") %>%
  #  read.csv(., row.names = 1)
  
  # right now samples are rows
  # feature + metadata are columns
  
  # 1. PCA
  plot_pca(m,
           this_ft_tbl,
           this_metadata,
           pca_dir)
  
  # 2. Paired t-tests
  # Deltas Mother - Baby | HUU vs HEU
  # Heatmap
  
  paired_ttest_res <- run_paired_ttests(m,
                                        this_ft_tbl,
                                        this_metadata,
                                        ttest_dir)
  
  plot_heatmap(paired_ttest_res,
             this_ft_tbl,
             this_metadata,
             m,
             heatmap_dir)
  
  # 3. Unpaired t-tests
  # 4. Fold-change
  # 5. Volcano
  # 6. Box plots
  
  for(who in c("Maternal", "Cord")){
    
    who_inx <- which(this_metadata$origin == who)
    
    who_ft_tbl <- this_ft_tbl[who_inx,] %>% t()
    
    who_meta <- this_metadata[who_inx,] 
    
    tt_df <- perform_ttest(peak_table = who_ft_tbl,
                           metadata = who_meta,
                           group_var = group_var,
                           group1 = group1,
                           group2 = group2,
                           prefix = glue::glue("{Sys.Date()}_{m}_{group1}_{group2}_{who}"),
                           output_dir = ttest_dir)
    
    fc_df <- perform_foldchange(peak_table = who_ft_tbl,
                                metadata = who_meta,
                                group_var = group_var,
                                group1 = group1,
                                group2 = group2,
                                prefix = glue::glue("{Sys.Date()}_{m}_{group1}_{group2}_{who}"),
                                output_dir = fc_dir)
    
    plot_volcano(m = m,
                 tt_res = tt_df,
                 pval_thresh = 0.05,
                 fc_res = fc_df,
                 fc_thresh = 1,
                 prefix = glue::glue("{Sys.Date()}_{m}_{group1}_{group2}_{who}"),
                 output_dir = volcano_dir,
                 pvalue_type = "adjusted",
                 width = 6,
                 height = 5,
                 anot_df = this_annots,
                 highlight_metabolites = TRUE,
                 metabolite_pattern = "efavirenz")
    
    volcano_df <- read.csv(glue::glue("{volcano_dir}{Sys.Date()}_{m}_{group1}_{group2}_{who}_volcano_df.csv")) 
    
    top_volcano <- volcano_df %>%
      filter(important != "No Change" & !is.na(annotation))
    
    top_volcano_ft <- top_volcano %>%
      filter(feature_id %in% colnames(this_ft_tbl)) %>%
      arrange(pvalue_adjusted) %>%
      group_by(annotation) %>%
      slice_head(n = 1) %>%
      ungroup() %>%
      arrange(pvalue_adjusted) %>%
      mutate(anot_level = gsub(".*_", "", annotation),
             keep = ifelse((anot_level %in% c("L1", "L2", "CSM") | !is.na(highlighted_met)), TRUE, FALSE)) %>%
      filter(keep)
    
    if(nrow(top_volcano_ft) > 20){
      top_volcano_ft <- top_volcano_ft[1:20,]
    }
    
    for(j in 1:nrow(top_volcano_ft)){
      
      this_top_volcano_ft <- top_volcano_ft[j,]$feature_id
      
      plot_boxplot_series(top_volcano_ft,
                        n_top = 5,
                        this_ft_tbl,
                        who_meta,
                        kp_df,
                        m,
                        who,
                        bp_dir,
                        analysis = "single",
                        this_feat = this_top_volcano_ft)
    }
    
    }
}

```

### Significantly different metabolites in HUU/HEU

```{r}

figure2_dir <- glue::glue("{results_dir}figure2/")
dir.create(figure2_dir)

# function to plot series

plot_series <- function(steroid_substrates_anots,
                        output_dir,
                        this_date,
                        met_data_dir,
                        cyp2b6_dir,
                        series_name){

  for(this_m in steroid_substrates_anots$mode %>% unique()){
    
    cat(this_m)
    
    this_met_output_dir <- glue::glue("{output_dir}{results_date}_{this_m}/")
    
    cat(this_met_output_dir)
    
    this_ft_tbl <- list.files(this_met_output_dir,
                              pattern = "feature_table",
                              full.names = TRUE) %>%
      read.csv(., row.names = 1) %>%
      t()
    
    this_metadata <- list.files(this_met_output_dir,
                                pattern = "metadata",
                                full.names = TRUE) %>%
      read.csv(., row.names = 1) %>%
      mutate(who = ifelse(origin == "Maternal", "Maternal", "Cord"),
             plot_group = paste0(who, "_", status))
    
    kp_path <- glue::glue("{met_data_dir}/12_03_2025_empirical_compounds/{this_m}_empirical_compounds.csv")
    
    kp_df <- read.csv(kp_path) %>%
      mutate(annotation = NA)
    # 
    # kp_df <- glue::glue("{this_met_output_dir}{this_m}_kp_df_anot.csv") %>%
    #   read.csv(., row.names = 1)
    
    rownames(kp_df) <- kp_df$id
    
    print(head(kp_df))
    
    these_steroids <- steroid_substrates_anots %>%
      filter(mode == this_m)
    
    for(this_who in unique(these_steroids$origin)){
      
      cat(this_who)
      
      these_feats = these_steroids %>%
        filter(origin == this_who) 
      
      who_inx <- which(this_metadata$origin == this_who)
      
      who_ft_tbl <- this_ft_tbl[who_inx,]
      who_meta <- this_metadata[who_inx,] 
      
      n_feats = nrow(these_feats)
      n_row = max(1, round(n_feats/5))
      plot_height = 2.5 * n_row
      plot_width = 1.4 * round(n_feats/n_row)
      
      plot_boxplot_series(these_feats,
                          n_top = n_feats,
                          who_ft_tbl,
                          who_meta,
                          kp_df,
                          this_m,
                          this_who,
                          cyp2b6_dir,
                          analysis = "series",
                          series_name = series_name,
                          nrow_facet = n_row,
                          height = plot_height,
                          width = plot_width)
      
      for(j in 1:nrow(these_feats)){
      
      this_top_ft <- these_feats[j,]$feature_id
      
      print(these_feats)
      
      plot_boxplot_series(these_feats,
                        n_top = 5,
                        who_ft_tbl,
                        who_meta,
                        kp_df,
                        this_m,
                        this_who,
                        cyp2b6_dir,
                        analysis = "single",
                        this_feat = this_top_ft)
    }
    }
    
  }
}

volcano_results_list <- vector("list")

volcano_results <- list.files(volcano_dir,
                              pattern = "volcano_df",
                              full.names = TRUE)

for(m in modes){
  
  this_met_output_dir <- glue::glue("/Volumes/T7/Li Lab/github/2025-08-04_HEU_analysis/immune_profile_analysis/heu_metabolomics_manuscript/{results_date}_processed_data/{results_date}_{m}")
  
  this_annots <- list.files(this_met_output_dir,
                            pattern = "annotations",
                            full.names = TRUE) %>%
    read.csv(., row.names = 1) %>%
    rename(feature_id = id_number)
  
  cat(nrow(this_annots), m)
  
  this_m_volcano_results <- grep(m, volcano_results)
  
  for(i in this_m_volcano_results){
    
    volcano_res <- read.csv(volcano_results[i], row.names = 1) %>%
      dplyr::left_join(., this_annots) %>%
      arrange(pvalue) %>%
      dplyr::filter(pvalue < 0.05) %>%
      mutate(file = basename(volcano_results[i]))
    
    volcano_results_list[[basename(volcano_results[i])]] <- volcano_res
    
  }
  
}

volcano_results_df <- volcano_results_list %>%
  dplyr::bind_rows() %>%
  dplyr::filter(pvalue_adjusted < 0.01) %>%  
  mutate(mode = case_when(
    grepl("RP_neg", file) ~ "RP_neg",
    grepl("RP_pos", file) ~ "RP_pos",
    grepl("HILIC_neg", file) ~ "HILIC_neg",
    grepl("HILIC_pos", file) ~ "HILIC_pos"),
    origin = ifelse(grepl("Cord", file), "Cord", "Maternal")
  )

```

### Steroid Precursors

```{r}

steroid_precursors_dir <- glue::glue("{figure2_dir}steroid_precursors/")
dir.create(steroid_precursors_dir)

steroid_precursors <- c("cholesterol", "pregnenolone")

steroid_precursors_anots <- volcano_results_df %>%
  filter(grepl(paste0(steroid_precursors, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(steroid_precursors_anots,
            output_dir,
            this_date,
            met_data_dir,
            steroid_precursors_dir,
            series_name = "Steroid Precursors")

```

### Bile acids

```{r}

bile_dir <- glue::glue("{figure2_dir}bile_acids/")
dir.create(bile_dir)

cholic_anots <- volcano_results_df %>%
  filter(grepl("cholic", tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(cholic_anots,
            output_dir,
            this_date,
            met_data_dir,
            bile_dir,
            series_name = "Bile Acids")
```

### Tryptophan + Derivatives

```{r}

trp_dir <- glue::glue("{figure2_dir}tryptophan/")
dir.create(trp_dir)

trp_substrates <- c("tryptophan", "kynur", "serotonin")

trp_substrates_anots <- volcano_results_df %>%
  filter(grepl(paste0(trp_substrates, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(trp_substrates_anots,
            output_dir,
            this_date,
            met_data_dir,
            trp_dir,
            series_name = "Tryptophan")

```

### PUFAs

```{r}

pufa_dir <- glue::glue("{figure2_dir}pufas/")
dir.create(pufa_dir)

pufa_substrates <- c("linoleic", "linolenic", 
                     "arachidonic", "aenoic", "adrenic")

pufa_substrates_anots <- volcano_results_df %>%
  filter(grepl(paste0(pufa_substrates, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(pufa_substrates_anots,
            output_dir,
            this_date,
            met_data_dir,
            pufa_dir,
            series_name = "PUFAs")
```

### Microbiome

```{r}

micro_dir <- glue::glue("{figure2_dir}microbiome/")
dir.create(micro_dir)

microbiome <- c(
  "phenylacetyl", "phenol sulfate",
  "cresol", "dopamine"
)

microbiome_anots <- volcano_results_df %>%
  filter(grepl(paste0(microbiome, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(microbiome_anots,
            output_dir,
            this_date,
            met_data_dir,
            micro_dir,
            series_name = "Microbiome")
```

### CYP2B6

```{r}

cyp2b6_dir <- glue::glue("{figure2_dir}CYP2B6_substrates/")
dir.create(cyp2b6_dir)

steroid_substrates <- c("testerone", "estradiol",
             "estrone", "cortisol", 
             "corticosterone", "aldosterone", "androsterone")

steroid_substrates_anots <- volcano_results_df %>%
  filter(grepl(paste0(steroid_substrates, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

pufa_substrates <- c("linoleic", "arachidonic",
                     "eicosapentaenoic")

pufa_substrates_anots <- volcano_results_df %>%
  filter(grepl(paste0(pufa_substrates, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(steroid_substrates_anots,
            output_dir,
            this_date,
            met_data_dir,
            cyp2b6_dir,
            series_name = "Steroids")

plot_series(pufa_substrates_anots,
            output_dir,
            this_date,
            met_data_dir,
            cyp2b6_dir,
            series_name = "PUFA")

```

### Sulfates

```{r}

sulf_dir <- glue::glue("{figure2_dir}sulfated_metabolites/")
dir.create(sulf_dir)

sulfate_substrates <- c("sulfate", "sulphate")

sulfate_substrates_anots <- volcano_results_df %>%
  filter(grepl(paste0(sulfate_substrates, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

# 1. Endogenous Steroids & Steroid-Derived
steroids <- c(
  "Estrone-3-sulfate_L2",
  "Equilin sulfate_L4",
  "5alpha-Pregnan-3beta,20alpha-diol disulfate_L4",
  "Androstane-3,7-diol disulfate_CSM",
  "Androstane-3,17-diol 17-sulfate_L4",
  "Pregnenolone-3,21-disulfate_CSM",
  "Pregnenolone sulfate_CSM",
  "Dehydroepiandrosterone sulfate; LC-tDDA; CE30_L2",
  "17-beta-estradiol 3-sulfate-17-(beta-D-glucuronide)_L4",
  "Estriol 3-sulfate 16-glucuronide_L4",
  "Estrone sulfate_CSM",
  "Corticosterone sulfate_CSM",
  "Cortisol 21-sulfate_L4",
  "Cholesterol sulfate_CSM",
  "Cholesterol sulfate_L4",
  "24-methylene-cholesterol sulfate_L4",
  "Vitamin D3 sulfate_L4",
  "25-Hydroxyvitamin D3 3-sulfate ester_L4",
  "24S-hydroxycholesterol 3-sulfate, 24-D-glucuronide_L4"
)

# 2. Bile Acids
bile_acids <- c(
  "Chenodeoxycholic acid sulfate_L4",
  "Chenodeoxycholic acid 3-sulfate_CSM"
)

# 3. Neurotransmitters & Neuroactive Metabolites
neuro <- c(
  "HOMOVANILLIC ACID SULFATE_L2",
  "Homovanillic acid sulfate_L4",
  "Dopamine 3-O-sulfate_CSM",
  "Epinephrine sulfate_L4",
  "Serotonin sulfate_L4"
)

# 4. Microbiome-Derived Metabolites
microbiome <- c(
  "P-CRESOL SULFATE_L2",
  "3-Indoxyl sulphate_L0",
  "Indole-3-carboxilic acid-O-sulphate_L4",
  "3-Hydroxyhippuric acid sulfate_L4",
  "D-3-Phenyllactic acid sulfate_L2",
  "5'-(3',4'-Dihydroxyphenyl)-gamma-valerolactone sulfate_L4"
)

# 5. Dietary Polyphenols / Flavonoids / Plant Metabolites
polyphenols <- c(
  "Isorhamnetin 3-glucuronide-7-sulfate_L4",
  "Daidzein 4'-sulfate_L4",
  "Vanillin-4-sulfate_L2",
  "3-Methoxyphenol sulfate_L2",
  "2-(4-Methoxyphenyl)ethyl hydrogen sulfate_L4",
  "Tyrosol 4-sulfate_L4",
  "Quercetin 3-sulfate-7-alpha-arabinopyranoside_L4",
  "5'-(3',4'-Dihydroxyphenyl)-gamma-valerolactone sulfate_L4",
  "Orientin 7-O-sulfate_L4",
  "Luteolin 4'-sulfate_L4",
  "8-Hydroxyluteolin 8-sulfate_L4",
  "6-Hydroxyluteolin 6,3'-dimethyl ether 7-sulfate_L4",
  "Apigenin 7-sulfate_L4",
  "Isorhamnetin 3-sulfate_L4",
  "Naringenin 5-sulfate_L4",
  "cis-Resveratrol 3-sulfate_L4",
  "Resveratrol 3-sulfate_CSM",
  "Vanillic acid 4-O-sulfate_L4",
  "3-Hydroxy-2-methyl-4-pyrone sulfate_L2",
  "Zearalenone 4-sulfate_L4",
  "Quercetin 7,3',4'-trimethyl ether 3-sulfate_L4",
  "Enterolactone 3''-sulfate_CSM"
)

# 6. Glycosaminoglycans & Carbohydrate Sulfates
glycans <- c(
  "Chondroitin sulfates_CSM",
  "Dermatan sulfate_L4",
  "chondroitin sulfate E (GalNAc4,6diS-GlcA), precursor 5a_L4",
  "N-Acetylgalactosamine 4-sulphate_CSM",
  "n-Acetylgalactosamine-6-sulfate_CSM",
  "Glucosamine 6-sulfate_L4"
)

# 7. Lipids & Sterol-like Compounds
lipids <- c(
  "beta-Sitosteryl sulfate_CSM",
  "13-chloro-docosane-1,14-disulfate_L4"
)

for(set in c("lipids", "glycans", "polyphenols",
             "microbiome", "neuro", "bile_acids",
             "steroids")){
  
  print(set)
  
  plot_series(sulfate_substrates_anots %>%
              filter(annotation %in% get(set)),
            output_dir,
            this_date,
            met_data_dir,
            sulf_dir,
            series_name = janitor::make_clean_names(set, case = "sentence"))
  
}

```

### Sugars

```{r}

sugar_dir <- glue::glue("{figure2_dir}sugars/")
dir.create(sugar_dir)

sugars <- c("glucose", "lactose", "fucose",
            "xylose")

sugars_anots <- volcano_results_df %>%
  filter(grepl(paste0(sugars, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(sugars_anots,
            output_dir,
            this_date,
            met_data_dir,
            sugar_dir,
            series_name = "Sugars")

```

### Heme

```{r}

heme_dir <- glue::glue("{figure2_dir}heme/")
dir.create(heme_dir)

heme <- c("biliverdin", "bilirubin", "stercobilin", "stercobilinogen",
          "urobilinogen")

heme_anots <- volcano_results_df %>%
  filter(grepl(paste0(heme, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(heme_anots,
            output_dir,
            this_date,
            met_data_dir,
            heme_dir,
            series_name = "Heme")


```

### Glycans

```{r}

glycan_dir <- glue::glue("{figure2_dir}glycans/")
dir.create(glycan_dir)

glycans <- c("acetylneuraminic acid", "glucosamine", "galactosamine")

glycans_anots <- volcano_results_df %>%
  filter(grepl(paste0(glycans, collapse = "|"), tolower(annotation))) %>%
  select(feature_id, 
         annotation,
         mode,
         origin)

plot_series(glycans_anots,
            output_dir,
            this_date,
            met_data_dir,
            glycan_dir,
            series_name = "Glycans")


```

### Export tables

```{r}

volcano_results <- list.files(volcano_dir,
                              pattern = "supp_volcano_table",
                              full.names = TRUE)

for(m in modes){
  
  this_met_output_dir <- glue::glue("{output_dir}{this_date}_{m}/")
  
  this_annots <- list.files(this_met_output_dir,
                            pattern = "annotations",
                            full.names = TRUE) %>%
    read.csv(., row.names = 1) %>%
    rename(feature_id = id_number)
  
  cat(nrow(this_annots), m)
  
  this_m_volcano_results <- grep(m, volcano_results)
  
  for(i in this_m_volcano_results){
    
    volcano_res <- read.csv(volcano_results[i], row.names = 1) %>%
      dplyr::left_join(., this_annots) %>%
      arrange(TT_adjpvalue) %>%
      write.csv(glue::glue("/Volumes/T7/Li Lab/Projects/HEU/2025-12-04_HEU_Metabolomics_Manuscript_V3/volcano_results/{basename(volcano_results[i])}"))
    
  }
  
}

volcano_results <- list.files(volcano_dir,
                              pattern = "volcano_df",
                              full.names = TRUE)

for(m in modes){

  this_m_volcano_results <- grep(m, volcano_results)
  
  all_kp <- glue::glue("{met_data_dir}/12_03_2025_empirical_compounds/{m}_empirical_compounds.csv") %>% 
    read.csv(., row.names = 1)
  
  for(i in this_m_volcano_results){
    
    volcano_res <- read.csv(volcano_results[i], row.names = 1)
    
    cat(nrow(volcano_res), m)
    
    volcano_res_mummichog <- volcano_res %>%
      select(feature_id, pvalue, statistic) %>%
      dplyr::left_join(.,
                       all_kp %>% tibble::rownames_to_column(., "feature_id")) %>%
      dplyr::select(mz, rtime, pvalue,
             statistic, feature_id) %>%
      dplyr::rename('m/z' = "mz",
                    'retention time' = "rtime",
                    "p-value" = "pvalue") %>%
      write_tsv(., 
                glue::glue("~/Desktop/JAX_Python/HEU_mummichog/{tools::file_path_sans_ext(basename(volcano_results[i]))}_mummichog_input.tsv"))
      
  }
  
}

```

## Efavirenz focus

### Efavirenz and metabolites boxplots and correlation

```{r}

efa_dir <- glue::glue("{figure2_dir}efavirenz/")
dir.create(efa_dir)

m = "RP_neg"

this_met_output_dir <- glue::glue("/Volumes/T7/Li Lab/github/2025-08-04_HEU_analysis/immune_profile_analysis/heu_metabolomics_manuscript/{results_date}_processed_data/{results_date}_{m}/")

this_ft_tbl <- list.files(this_met_output_dir,
                          pattern = "feature_table",
                          full.names = TRUE) %>%
  read.csv(., row.names = 1) %>%
  t()

this_metadata <- list.files(this_met_output_dir,
                            pattern = "metadata",
                            full.names = TRUE) %>%
  read.csv(., row.names = 1) %>%
  mutate(who = ifelse(origin == "Maternal", "Maternal", "Cord"),
         plot_group = paste0(who, "_", status))

this_annots <- list.files(this_met_output_dir,
                          pattern = "annotations",
                          full.names = TRUE) %>%
  read.csv(., row.names = 1) %>%
  rename(feature_id = id_number)

kp_df <- glue::glue("{this_met_output_dir}{m}_kp_df_anot.csv") %>%
  read.csv(., row.names = 1)

# Efavirenz_L2

ef_df_list <- vector("list")

for(who in c("Maternal", "Cord")){
  
  volcano_df <- read.csv(glue::glue("{volcano_dir}{Sys.Date()}_{m}_{group1}_{group2}_{who}_volcano_df.csv")) 
  
  volcano_df_efa <- volcano_df %>%
    filter(annotation == "Efavirenz_L2") %>%
    dplyr::left_join(., kp_df,
                     by = join_by(feature_id == id))
  
  for(this_feat in volcano_df_efa$feature_id){
    
    plot_boxplot_series(volcano_df,
                        n_top = 5,
                        this_ft_tbl,
                        this_metadata %>% filter(origin == who),
                        kp_df,
                        m,
                        who,
                        efa_dir,
                        "single",
                        this_feat)
  }
  
  # 8-Hydroxyefavirenz_L4
  
  volcano_df_8efa <- volcano_df %>%
    filter(annotation == "8-Hydroxyefavirenz_L4") %>%
    dplyr::left_join(., kp_df,
                     by = join_by(feature_id == id))
  
  for(this_feat in volcano_df_8efa$feature_id){
    
    plot_boxplot_series(volcano_df,
                        n_top = 5,
                        this_ft_tbl,
                        this_metadata %>% filter(origin == who),
                        kp_df,
                        m,
                        who,
                        efa_dir,
                        "single",
                        this_feat)
  }
  
  # Correlation between the 2
  
  # x = efavirenz F29911
  # y = 8-hydroxy F21642 
  
  efa_ft <- "F29911"
  hyroxy_ft <- "F21642" 
  
  who_inx <- which(this_metadata$origin == who)
    
  who_ft_tbl <- this_ft_tbl[who_inx,] 
    
  who_meta <- this_metadata[who_inx,] 
  
  ef_df <- who_ft_tbl %>%
    as.data.frame() %>%
    select(efa_ft, hyroxy_ft)
  
  ef_df_list[[who]] <- ef_df
  
  efa_ft_info <- kp_df %>%
    filter(id == efa_ft)
  
  hydroxy_ft_info <- kp_df %>%
    filter(id == hyroxy_ft)
  
  efa_cor <- cor.test(ef_df[[efa_ft]], ef_df[[hyroxy_ft]])
  
  p_cor_ef <- ggplot2::ggplot(data = ef_df,
                              aes(x = .data[[efa_ft]],
                                  y = .data[[hyroxy_ft]])) +
    geom_point(size = 3,
               fill = "#D41159",
               color = "black",
               alpha = 0.75,
               shape = 21, ) +
    geom_smooth(method='lm',
                se = FALSE,
                color = "black") +
    xlab(glue::glue("Efavirenz\n{efa_ft} - {round(efa_ft_info$mz, digits=4)} m/z @ {efa_ft_info$rtime}s")) +
    ylab(glue::glue("{hydroxy_ft_info %>% pull(annotation)}\n{hyroxy_ft} - {round(hydroxy_ft_info$mz, digits=4)} m/z @ {hydroxy_ft_info$rtime}s")) +
    theme_classic() +
    labs(
      title = glue::glue("ρ in {who}: {round(efa_cor$estimate, digits = 2)}")
    ) +
    theme(
      text = element_text(family = "Arial"),
      strip.background = element_blank(),
      axis.text = element_text(size = 9),
      axis.title.x = element_text(size = 9.5, margin = margin(t = 1.5)), # reduce space above x-axis title
      axis.title.y = element_text(size = 9.5, margin = margin(r = 1.5)), # reduce space right of y-axis title
      plot.title = element_text(size = 11, hjust = 0.5)) 
  
  ggsave(filename = glue::glue("{efa_dir}{Sys.Date()}_{m}_{who}_efa2efa_{efa_ft}_{hyroxy_ft}_corr_plot.png"),
         width = 2.7,
         height = 3)
  
}

```

### Correlation between mother and baby

```{r}

mom_efa <- ef_df_list$Maternal %>%
  rename_with(~ paste0(., "_M")) %>%
  tibble::rownames_to_column(., "sample_name") %>%
  dplyr::left_join(., this_metadata %>%
                     mutate(dyad_id = gsub("_.*", "", Sample_ID)) %>%
                     select(sample_name, dyad_id))

cord_efa <- ef_df_list$Cord %>%
  rename_with(~ paste0(., "_C")) %>%
  tibble::rownames_to_column(., "sample_name") %>%
  dplyr::left_join(., this_metadata %>%
                     mutate(dyad_id = gsub("_.*", "", Sample_ID)) %>%
                     select(sample_name, dyad_id)) %>%
  select(-sample_name) %>%
  dplyr::left_join(., mom_efa %>% select(-sample_name)) %>%
  dplyr::filter(complete.cases(.))

for(this_feat in c(hyroxy_ft, efa_ft)){
  
  momft <- paste0(this_feat, "_M")
  cordft <- paste0(this_feat, "_C")
  
  efa_cor_mb <- cor.test(cord_efa[[momft]], cord_efa[[cordft]])
  
  ft_info <- kp_df %>%
    filter(id == this_feat) 
  
  p_cor_ef_mb <- ggplot2::ggplot(data = cord_efa,
                              aes(x = .data[[momft]],
                                  y = .data[[cordft]])) +
    geom_point(size = 3,
               fill = "#D41159",
               color = "black",
               alpha = 0.75,
               shape = 21, ) +
    geom_smooth(method='lm',
                se = FALSE,
                color = "black") +
    xlab(glue::glue("Maternal")) +
    ylab(glue::glue("Cord")) +
    theme_classic() +
    labs(
      title = glue::glue("ρ: {round(efa_cor_mb$estimate, digits = 2)}\n{ft_info$annotation}\n{ft_info$mz} @ {round(ft_info$rtime, 2)}")
    ) +
    theme(
      text = element_text(family = "Arial"),
      strip.background = element_blank(),
      axis.text = element_text(size = 9),
      axis.title.x = element_text(size = 9.5, margin = margin(t = 1.5)), # reduce space above x-axis title
      axis.title.y = element_text(size = 9.5, margin = margin(r = 1.5)), # reduce space right of y-axis title
      plot.title = element_text(size = 11, hjust = 0.5)) 
  
  ggsave(filename = glue::glue("{efa_dir}{Sys.Date()}_{m}_{this_feat}_maternal_cord_corr_plot.png"),
         width = 2.7,
         height = 3)
  
}

```

### Efavirenz correlation to all other features

```{r}

efa_cors <- vector("list")

for(this_who in c("Maternal", "Cord")){
  
  this_ef_df <- this_ft_tbl %>%
    as.data.frame() %>%
    select(efa_ft, hyroxy_ft) %>%
    tibble::rownames_to_column(.) %>%
    dplyr::rename(sample_name = rowname) %>%
    filter(sample_name %in% 
           (this_metadata %>% 
           filter(origin == this_who) %>% 
           pull(sample_name))) %>%
    select(F29911) 
  
  other_ft_df <- this_ft_tbl %>%
    as.data.frame() %>%
    tibble::rownames_to_column(.) %>%
    dplyr::rename(sample_name = rowname) %>%
    filter(sample_name %in% 
             (this_metadata %>% 
                filter(origin == this_who) %>% 
                pull(sample_name))) %>%
    select(-c(efa_ft, sample_name)) 
  
  # correlations (vector)
  cors <- cor(this_ef_df, other_ft_df, use="pairwise.complete.obs")
  
  # sample size (pairwise complete obs)
  n <- apply(cbind(this_ef_df, other_ft_df), 2, function(col) sum(complete.cases(this_ef_df, col)))[-1]
  
  # t-statistics
  tvals <- cors * sqrt( (n - 2) / (1 - cors^2) )
  
  # two-sided p-values
  pvals <- 2 * pt(-abs(tvals), df = n - 2)
  
  # corrected p-values
  padj <- p.adjust(pvals, method = "BH")
  
  # results table
  cor_results <- data.frame(
    feature_id = colnames(cors),
    correlation = cors[1,],
    pvalue = pvals[1,],
    pvalue_adjusted = padj
  ) 
  
  volcano_df <- read.csv(glue::glue("{volcano_dir}{Sys.Date()}_{m}_{group1}_{group2}_{this_who}_volcano_df.csv"))

  all_kp <- glue::glue("{met_data_dir}/12_03_2025_empirical_compounds/{m}_empirical_compounds.csv") %>%
    read.csv(., row.names = 1)

  both_df <- dplyr::bind_rows(cor_results %>%
                                select(feature_id, pvalue_adjusted) %>%
                                mutate(type = "Features correlated with Efavirenz"),
                              volcano_df %>%
                                select(feature_id, pvalue_adjusted) %>%
                                mutate(type = "Features significantly different between HUU/HEU")) %>%
    mutate(significant = pvalue_adjusted < 0.01) %>%
    group_by(feature_id) %>%
    mutate(count = sum(significant, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(shape = (count == 2)) %>%
    dplyr::left_join(., all_kp %>%
                       mutate(feature_id = rownames(all_kp)) %>%
                       select(feature_id, mz, rtime))

  efa_cors[[this_who]] <- both_df
  
  cat((sum(both_df$shape)/2), "features out of", nrow(cor_results), "found in both.")
  
  p_efz <- ggplot(both_df %>%
                    filter(rtime < 450),
                  aes(y = -log10(pvalue_adjusted),
                      x = rtime)) +
    geom_point(aes(fill = significant),
               shape = 21,
               color = "black",
               stroke = 0.5,
               size = 1.5,
               alpha = 0.7) +
    theme_minimal() +
    scale_fill_manual(values = c("TRUE" = "#D41159",
                                 "FALSE" = "darkgrey"),
                      guide = "none") +
    theme(
      text = element_text(family = "Arial"),
      axis.text = element_text(size = 9),
      axis.title.x = element_text(size = 9, margin = margin(t = 2.5)), # reduce space above x-axis title
      axis.title.y = element_text(size = 9, margin = margin(r = 2.5)), # reduce space right of y-axis title
      plot.title = element_text(size = 9.5, hjust = 0.5),
      strip.text = element_text(size = 10, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(
        fill = "grey90",     # background color
        color = "black",     # border color
        linewidth = 0.5      # border thickness
      )) +
    facet_wrap(. ~ type, nrow = 2) +
    labs(
      x = "Retention Time (seconds)",
      y = "-log10 Unadjusted p-value",
      shape = "Found in both",
      title = glue::glue("{this_who} ρ BH < 0.01")
    ) +
    geom_hline(yintercept = -log10(0.01),
               linetype = "dashed")

  ggplot2::ggsave(glue::glue("{efa_dir}{Sys.Date()}_{m}_{this_who}_efavirenz_all_features_correlations.png"),
                  width = 6,
                  height = 3.5)
  
  # Save outputs for mummichog
  cor_results_mummichog <- cor_results %>%
    dplyr::left_join(., all_kp %>%
                       mutate(feature_id = rownames(all_kp)) %>%
                       select(feature_id, mz, rtime)) %>%
    select(mz, rtime, pvalue, correlation, feature_id) %>%
    rename('m/z' = "mz",
           "retention time" = "rtime",
           "p-value" = "pvalue",
           "statistic" = "correlation") %>%
      write_tsv(., 
                glue::glue("~/Desktop/JAX_Python/HEU_mummichog/{Sys.Date()}_{m}_{this_who}_efavirenz_cor_mummichog_input.tsv"))
  
}
```

### Upset Plot Cord + Maternal

```{r}

# Prepare data for the UpSet plot
upset_df <- list(Efavirenz_Correlated_M = efa_cors$Maternal %>%
                   dplyr::filter(significant & 
                                   type == "Features correlated with Efavirenz") %>%
                   pull(feature_id),
                 HEU_Significant_M = efa_cors$Maternal %>%
                   dplyr::filter(significant & 
                                   type == "Features significantly different between HUU/HEU") %>%
                   pull(feature_id),
                 Efavirenz_Correlated_C = efa_cors$Cord %>%
                   dplyr::filter(significant & 
                                   type == "Features correlated with Efavirenz") %>%
                   pull(feature_id),
                 HEU_Significant_C = efa_cors$Cord %>%
                   dplyr::filter(significant & 
                                   type == "Features significantly different between HUU/HEU") %>%
                   pull(feature_id))

upset_m = make_comb_mat(upset_df, mode = "intersect")
upset_m_filt <- upset_m[comb_degree(upset_m) > 1]

png(glue::glue("{efa_dir}{Sys.Date()}_{m}_maternal_cord_upset_plot_filt.png"), 
    width = 1400, 
    height = 800, 
    res = 320)

UpSet(upset_m_filt,
      row_gap = unit(0, "mm"),
      set_order = c("Efavirenz_Correlated_C", "HEU_Significant_C",
                    "Efavirenz_Correlated_M", "HEU_Significant_M"),
      comb_col = c("#11224E", "#D41159",
                   "#CCA942", "#0C7D6B")[comb_degree(upset_m_filt)],
      
      top_annotation = upset_top_annotation(m = upset_m_filt,
                                            add_numbers = TRUE,
                                            annotation_name_gp = gpar(fontsize = 10, fontfamily = "Arial"),
                                            number_gp = gpar(fontsize = 10,
                                                             fontfamily = "Arial")),
      
      right_annotation = 
        upset_right_annotation(m = upset_m_filt,
                               add_numbers = TRUE,
                               annotation_name_gp = gpar(fontsize = 10, fontfamily = "Arial"),
                               number_gp = gpar(fontsize = 10, fontfamily = "Arial")),
      
      # main matrix text (row/column names)
      row_names_gp = gpar(fontsize = 10, fontfamily = "Arial"),
      column_names_gp = gpar(fontsize = 10, fontfamily = "Arial"))

dev.off()

```

```{r}

efv_cor_dir <- "/Users/jasminechong/Desktop/JAX_Python/HEU_mummichog/"

cord <- glue::glue("{efv_cor_dir}2025-12-16_HILIC_neg_Cord_efavirenz_cor_mummichog_input.tsv") %>%
  readr::read_delim(.) %>%
  mutate(padj = p.adjust(as.numeric(`p-value`), method = "BH"))

cord_sig <- cord %>%
  filter(padj < 0.01) %>%
  nrow()

maternal <- glue::glue("{efv_cor_dir}2025-12-16_HILIC_neg_Maternal_efavirenz_cor_mummichog_input.tsv") %>%
  readr::read_delim(.) %>%
  mutate(padj = p.adjust(as.numeric(`p-value`), method = "BH"))

maternal_sig <- maternal %>%
  filter(padj < 0.01) %>%
  nrow()

```

## Mummichog

```{r}

mum_output_dir <- glue::glue("{figure2_dir}mummichog/")
dir.create(mum_output_dir)

# For HILIC Pos
mum_results_path <- "/Volumes/T7/Li Lab/Projects/HEU/OLD/2025-12-14_HEU_Metabolomics_V3/heu_data/SL_mummichog/"

# OLD results 
sl_path <- "/Volumes/T7/Li Lab/Projects/HEU/OLD/2025-10-01_HEU_datafreeze/metabolomics/SL-20250914/"

mummichog_results <- list.files(path = sl_path,
                                recursive = TRUE,
                                pattern = "mcg_pathwayanalysis_mcg_.*\\.xlsx$",
                                full.names = TRUE) 

mummichog_results <- mummichog_results[str_detect(mummichog_results, "RP_neg")] %>%
  # map_dfr(~ readr::read_delim(.x, show_col_types = FALSE) %>%
  #           mutate(source_file = basename(.x))) %>%
  map_dfr(~ readxl::read_excel(.x) %>%
          mutate(source_file = basename(.x))) %>%
  dplyr::bind_rows() %>%
  dplyr::rename(pvalue = 'p-value') %>%
  group_by(source_file) %>%  
  mutate(p_adj = p.adjust(pvalue, method = "BH")) %>%
  ungroup() 

# HILIC results only
m = "RP_neg"

hilic_mummichog_results <- mummichog_results %>%
  select(pathway, pvalue, p_adj, source_file) %>%
  filter(str_detect(source_file, "RP")) %>%
  mutate(test = ifelse(grepl("Efavirenz", source_file), "EFV_Cor", "HEU_HUU"),
         who = ifelse(grepl("mothers", source_file), "Maternal", "Cord"),
         group = paste0(who, "\n", test),
         group = factor(group,
                        levels = c("Cord\nEFV_Cor", "Cord\nHEU_HUU", 
                                   "Maternal\nEFV_Cor", "Maternal\nHEU_HUU"))) %>%
  group_by(pathway) %>%
  filter(any(pvalue < 0.05, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(logq = -log10(pvalue),
         is_sig = factor(pvalue < 0.05),
         pathway = factor(pathway, levels = sort(unique(pathway), decreasing = TRUE)))

# Define shapes and colors
shapes <- c("FALSE" = 1, "TRUE" = 16)
colors <- c("FALSE" = "darkgrey", "TRUE" = "black")

# Plot
p <- ggplot(hilic_mummichog_results,
            aes(x = group, y = pathway)) +
  geom_point(aes(size = logq, shape = is_sig, color = is_sig)) +  # separate points horizontally for the same pathway
  scale_shape_manual(values = shapes) +
  scale_color_manual(values = colors) +
  scale_size_continuous(range = c(2,6)) +
  theme_minimal() +
  guides(
    color = "none",
    shape = "none"
  ) +
  labs(x = "",
       y = "Pathway",
       size = "-log10 P-value",
       shape = "",
       color = "") +
  theme(
    text = element_text(family = "Arial"),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 9),
    axis.title = element_text(size = 10.5),
    axis.line = element_line(color = "black"))

ggplot2::ggsave(glue::glue("{mum_output_dir}{Sys.Date()}_{m}_mummichog_results_plot_rawp.png"),
                width = 8,
                height = 7)

```
